---
title: "Rlpj Package"
author: "Ramiro Silveyra Gonzalez, Maurizio Bagnara, Florian Hartig"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Rlpj Package}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

The Rlpj package is thought to ease the parallelization of the LPJ-GUESS in High Computer Performance Environments (HPC). The package is a collection of tools that will allow you to create a parallel setup and run in parallel not only the LPJ model, but also the processing of the model outputs. The package can be also used for parallel runs on personal computers.

In this document you will learn how to prepare a parallel setup and how to use the different functions of Rlpj. The code is open to further improvements and modifications, feel free to adapt it to your needs.

## Overview

#### General aspects

The Rlpj package will help you to run LPJ-GUESS in parallel,  but first you will have to prepare your data, so that you can make use the of the package utilities. Specifically, Rlpj requires that *i)* you have already compiled the LPJ_GUESS model in your computer or home folder in the HPC, and that *ii)* the existence of a directory where the outputs of the model will be saved, lets name it the Main Directory. 


In the Main Directory, the following files must be available:

  - a link to the guess executable 
  - a text file with the grids 
  - the model input templates (optional)

with a structure like this:

  $ cd ./mainDir
  
  $ ls   
  
          guess / guesscmd.exe   # link to the model executable (not for Windows)
          gridlist.txt           # list of grid cells
          global.ins             # model template  (optional)
          global_cru.ins         # model template  (optional) or
          global_cf.ins          # model template  (optional)



The model templates are included within the package as part of the system data. They have been edited with some reference values, but you might want to change then. The function *getTemplate* allows to obtain the templates from package. To learn more about how to writing parameters in the template, check out the *writeParameters* function and the *getParameterList*. 


#### Running LPJ in parallel

Running the LPJ parallel involves two steps. First, to create a parallel setup (**setupParallel**) function, and second, to actually run in parallel the model (**runLPJ**).  The parallelization requires the package *snow* for SOCK cluster and for MPI cluster, the *Rmpi* package.

The **setupParallel** function will help you at creating a directory structure for storing your model outputs and arranging the model templates for each run. Calling the setupParallel returns a list object that contains all information needed to run the LPJ in parallel.

The **runLPJ** function reads the setup parallel object and creates a cluster to which submits the model call with its respective parameters. 



## An Example of using Rlpj

Here you will see an example of how to use the Rlpj package. We will not run the LPJ-GUESS itself because the model is not included in the package. However, we will make use of *almost* all package functions.


#### First steps


Lets choose a mainDir and have a look at it. Remember that the mainDir should contain a link to your guess executable and file with list of grid cells.

```{r}
# clearing local environment
rm(list=ls())
# Loading the library
library(Rlpj)

# The main dir in this case is within the package
mainDir <- system.file("extdata", package = "Rlpj")
list.files(mainDir) 

```

Now we should first specicify paths to input data and setting options for runLPJ.

```{r, fig.show='hold'}
# We need to specify the absolute path of each input file :
file_co2 <-"/some/absolute/path/crudata/co2_1901-2006.txt"
file_cru <- "/some/absolute/path/crudata/cru_1901_2006.bin"
file_cru_misc <- "/some/absolute/path/crudata/cru_1901_2006misc.bin"
file_ndep <- "/some/absolute/path/crudata/GlobalNitrogenDeposition.bin"

# If we are using the global_cf.ins file you need to specify the site specific input files as well
file_temp <- "/some/absolute/path/cfdata/temp.nc"
file_prec <- "/some/absolute/path/cfdata/prec.nc"
file_insol <- "/some/absolute/path/cfdata/rad.nc"

# The gridlist should be place in the  mainDir
gridList <- "gridlist.txt"

settings <- list (gridList = gridList,mode = "cf", scale = "global",
                  file.co2 = file_co2, file.cru = file_cru, file.cru.misc = file_cru_misc,
                  file.ndep = file_ndep, file.temp = file_temp, file.prec = file_prec,
                  file_insol = file_insol)

```

#### A serial run

After the steps above, we are ready to call runLPJ. LPJ-GUESS is, however, not part of this 
package, and thus, we will skip runLPJ here. Anyway, we should procceed as follows

```{r}


# ---- The call ------------------------#
#result <-   runLPJ(mainDir,  settings= settings)
# ---- The call ------------------------#

```
The above call will run the model wiht the default values for global_cf template. We can specify other values 
for the model parameters and also choose what model output we want to create. In the next section, we will do when 
using the package for parallel runs of runLPJ.

The output is a LPJData object holding the information on the model run and the data outputs

    result
  
      class              : LPJData
      
      LPJ template 1     : global.ins
      
      LPJ template 2     : global_cf.ins
      
      grid cells         : ! grid_ID  Location(spaces are ok) 0  Somewhere
      
      run directory      : /some/absolute/paht/runDirectory
      
      LPJ model outputs  : 39 outputs
      
      aaet agpp aiso amon anpp cflux clitter cmass cpool cton_leaf dens      
      firert fpc speciesheight lai maet mevap mgpp mintercep miso mlai mmon
      mnee mnpp mpet mra mrh mrunoff mwcont_lower mwcont_upper nflux ngases
      nlitter nmass npool nsources nuptake runoff vmaxnlim


#### A parallel run

Because we are  running this on a personal laptop, we will choose a low number cores and create a SOCK cluster. By calling the function setupParallel we will create a parallel setup for runLPJ. 

```{r}

# Lets call the setupParallel
options(error=traceback)
mySetup <- setupLPJParallel(numCores= 3, clusterType = "SOCK", mainDir=mainDir)
mySetup
```

If we check the mainDir, we see that there new folders. Within the the runDir folders, there is also a outDir folder to temporary store the model outputs. 

```{r}
list.files(mainDir) 
runDir <- file.path(mainDir, "runDirectory1")
list.files(runDir)
```

Now we can define the model outputs and the parameters to be tested. In this example, we will test several values for 
*emax* for europe. To see what are the default values for europe, we can use the *getParameterList* function.

```{r}
# Choosing the parameters
# Only modifying 1 parameter: emax. 
# The complete list of parameters that can be modified can be found in the "data_raw/createInternalData.R"
# Chekc parameterList
parameterDefault <- getParameterList("europe")
parameterDefault[1:10]

parameterDefault$run_emax

parameterDefault <- list (run_emax = NULL)
# I want to test 20 different values for emax. I want therefore to run 20 time the LPJ
par <- seq(1,5, len = 20)
print (par)
# I create the list object with the parameters
parameterList <- vector("list", length(par))
for (i in 1:length(par)) {
  parameterDefault$run_emax <- par[i]
  parameterList[[i]] <- parameterDefault
}
print(parameterList[c(2:3)])


# Now, I specify which outputs I want process.
typeList <- c("aaet", "nuptake")

```



Now we are ready to call the *runLPJ* . Because LPJ-GUESS is not part of this 
package, we will skip this step. The call to runLPJ would look like the one below.


```{r, fig.show='hold'}
# ---- The call ------------------------#
#result <-   runLPJ(mySetup,  settings= settings, parameterList = parameterList )
# ---- The call ------------------------#

```

The runLPJ  would return something like this:

    Output typeList has not been provided

    Setting typeList to default values
    

    Using package template (template 1)

    Saving package template in the mainDir
    

    Using package template (template 2)

    Saving package template in the mainDir
    

    Checking conditions

    Reading the parallel object structure

    |==============================================================| 100%

    Creating a SOCK cluster with 3 cores

    Sending tasks to the cores


    Processing ended!

 The output of the runLPJ call is a list  with the following structure:
 
    str(result,1)

    List of 6

    $ :Formal class 'LPJData' [package "Rlpj"] with 2 slots

    $ :Formal class 'LPJData' [package "Rlpj"] with 2 slots

    $ :Formal class 'LPJData' [package "Rlpj"] with 2 slots

    $ :Formal class 'LPJData' [package "Rlpj"] with 2 slots

    $ :Formal class 'LPJData' [package "Rlpj"] with 2 slots

    $ :Formal class 'LPJData' [package "Rlpj"] with 2 slots




## How Rlpj works

Here you will see how a single run takes place. The *runLPJ* does some preliminary taks such as organizing input data into the runDirs. If parallel, it creates the desired cluster. In both cases, runLPJ calls a internal wrapper which agglutinates all single steps to carry one complete call to LPJ-GUESS. Here we will run the steps that take in place in a single run.

```{r, fig.show='hold'}
# Lets move to one of the run directories
runDir <- file.path(mainDir, "runDirectory1")
list.files(runDir)
# First step is to write the parameters
# We will selected only one of the parameters created above
parameters <- parameterList[[1]]
# this we want to write in the template
parameters
# Now we call the writeTemplate function
# First we read the template to check how it looks like
# in concrete, the parameters we want to change
template <- getTemplate("europe", runDir)
list.files(runDir)

template <- readLines(file.path(runDir, "europe.ins"))

template [c(156:166)]
# now we write the template
writeTemplate(template1 = "europe.ins", parameterList = parameters, runDir = runDir)
template <- readLines(file.path(runDir, "europe.ins"))

# and lets check what it has been written in the template
# if no value was provided for the parameters, they have been set to default values
template [c(156:166)]

# Next step is to call the model
# but we skip it
# runLPJ(mainDir, runDir, "global_cru.ins")
```

Assuming that we would have run the LPJ, we should see in the outDirectory the output files, as specified in the template. In this example, we do not run the LPJ and thus, we do not create any output data. But some output data examples are provided in the folder *extdata/exampleOutputs*.  In the lines below, we will read the outputs, process and plot them.

```{r, fig.show='hold', fig.width=7, fig.heigth=10}
# Lets move to the ouput examples
outDir <- system.file("extdata/exampleOutputs", package = "Rlpj")
# And check the outputs
list.files(outDir)

# Lets say we want to process just some of the datatypes
# Normally, you will want to proccess all outputs you have created.
typeList <- c("aaet","lai", "nonsense")
# As you noticed, we included a wrong value: nonsense. The function should be able to sort this out.
# We call the getLPJData function to obtain the processed data
LPJout <- getLPJData(x = outDir,typeList = typeList,  runInfo=list(runNumber = 1, template1 = "europe.ins", template2="europe_cf.ins", runDir = "runDirectory1"))
# You see the fucntion has foudn we gave wrong values
# The returned object is a S4 Object
LPJout
slotNames(LPJout)
names(LPJout@dataTypes)
summary(LPJout@dataTypes$lai)
```

Now that we have processed the output data, we can plot the data and see how the variables evolve across the years. The *plotLPJData* function will plot the data as a time series using the zoo library. 

```{r,fig.show= "hold", fig.keep ='all',  fig.width=7, fig.heigth=16}
# Now we can plot the object
# Let say I just want to plot LAI, and I dont want to save the plots
plotLPJData(x = LPJout, typeList = c("lai") , outDir = outDir, save.plots = FALSE)
```

If you have more doubts about the functions, please check the help files.

```{r, echo=FALSE}
# And cleand the folder
files <- list.files(mainDir)
files <- files[!grepl("grid", files)]
files <- files[!grepl("example", files)]
unlink(file.path(mainDir, files), recursive = TRUE)
#do.call(file.remove,list( grep("runDirectory",list.files(mainDir) )) )
```
  
  
